Language learning via recursive pragmatic reasoning

### 2013-05-31 praglearn.tex
trivial changes

As soon as we attempt to add uncertainty about the meanings of words
to such a model, new paradoxes arise. Suppose the listener is a young
child who is uncertain about the lexicon their partner is using. The
obvious solution is for them to place a prior on the lexicon; they
then update their posterior based on whatever utterances and
contextual cues they observe, and in the mean time interpret each
utterance by making their best guess, marginalizing out this
uncertainty. This basic structure is captured in previous models of
Bayesian word learning \cite{frank2009}. But when combined with the
recursive pragmatic model, a new question arises: Given such a
listener, what model should the speaker use? A rational speaker
attempts to maximize the listener's likelihood of understanding, so if
an uncertain listener interpets by marginalizing over some posterior,
then a knowledgeable speaker should calculate this posterior and
marginalize themselves. But then the speaker is not using any
particular lexicon at all, and a rational learning listener should
therefore not be trying to infer the speaker's lexicon.\footnote{Of
  course, in reality both parties will generally have some
  uncertainty, making the situation even worse. If we start from an
  uncertain listener with a prior over lexicons, then a first-level
  uncertain speaker needs a prior over priors on lexicons, a
  second-level uncertain listener needs a prior over priors over
  priors, etc. The original $L_0 \rightarrow S_1 \rightarrow \dots$
  recursion was bad enough, but at least each step had a constant
  cost. This new recursion produces hyper$^n$-distributions for which
  inference almost immediately becomes intractable even in principle,
  since the dimensionality of the learning problem increases with each
  step. Yet, without this addition of new uncertainty at each level,
  the model would dissolve back into certainty as in the previous
  paragraph, making learning
  impossible.}
% In fact, such an uncertain listener's uncertainty dissolves; they
% may not know the lexicon, but they do
% know their posterior over the lexicon -- and because
% this is all that the speaker cares about, there is
% nothing for the listener to learn from the speaker's
% behavior.

@@

Formally, let an unadorned $L$ and $S$ denote our convention-based
listener and speaker. By assumption (c), the top-level listener has uncertainty over the lexicon; if the lexicon were known they would draw inferences as in $L_n$ above:
%\begin{align*}
%  P_L(\obj | \word) &= P_{S_{n-1}}(\word | \obj, \text{$L$'s data}) \times \prior \\
%    &= \sum_\lex P_{S_{n-1}}(\word | \obj, \lex) \times P(\lex | \text{$L$'s data}) \times \prior. \\
%    &= P_{L_n}(\obj | \word, \text{$L$'s data}). \\
%  P_S(\word | \obj) &\propto \exp\Big(\lambda \big(\log P_{L_n}(\obj | \word, \text{$S$'s data}) - \text{cost}(\word)\big)\Big).
%\end{align*}
%
\begin{align*}
  P_L(\obj | \word, \text{$L$'s data}) &=  \sum_\lex P_{L_n}(\obj | \word, \lex) P(\lex | \text{$L$'s data}) \\
\end{align*}
Here $L$'s data consists of $S$'s utterances together with whatever
other information $L$ may have about their meaning (e.g.~from
contextual clues). 
By assumption (b), $L$ treats these  utterances
as samples from the knowledgeable speaker $S_{n-1}$, not $S$, and thus
as being informative about the lexicon. For instance, when the data is a set of fully observed 
word-referent pairs $\{w_i, o_i\}$:
\begin{align*}
  P(\lex | \text{$L$'s data}) \propto P(\lex) \prod_i P_{S_{n-1}}(w_i|o_i,  \lex)
\end{align*}
The top-level speaker attempts to select the word which soft-maximizes their utility, with utility now being defined in terms of the informativity of the expectation (over lexicons) that the listener will have for at the right referent\footnote{An alternative model would have the speaker take the expectation over informativity, instead of the informativity of the expectation.}:
\begin{align*}
  P_S(\word | \obj, \text{$S$'s data}) &\propto \exp\Big(\lambda \big(\log  \sum_\lex P_{L_{n-1}}(\obj | \word, \lex) P(\lex|\text{$S$'s data}) - \text{cost}(\word)\big)\Big)
\end{align*}
Here $P(\lex|\text{$S$'s data})$ is defined similarly, when $S$ observes
$L$'s interpretations of various utterances, and treats them as
samples from $L_n$, not $L$.

@@

Many sets of words in natural language form scales in which each term
is more informative than, and is entailed by, the previous. ``Some'' and ``all'' form a scale of this
type. While ``I ate some of the cookies'' is compatible with the
followup ``in fact, I ate {\it all} of the cookies,'' the reverse is
not true. ``Might'' and ``must'' are another example, as are ``OK,''
``good,'' and ``excellent.'' All of these scales allow for {\it scalar
  implicatures} \cite{grice1975}: if a speaker uses a weaker term, a
listener is licensed to infer that the stronger term was not true in
this situation. So although ``I ate some of the cookies'' could in
principle be compatible with eating {\sc all} of them, the listener is lead
to believe that {\sc some-but-not-all}' is the likely state of
affairs. We use the term {\it specificity implicature} here (following
\cite{bergen2012}) to denote that we are agnostic about whether such
implicatures are conventionalized or whether they use a
contextually-defined scale.

The recursive pragmatic reasoning portions of our model capture
findings on specificity implicature. Consider a representation of
scalar implicature as a matrix $\left[
    \begin{array}{cc}
      0.5 & 0.5 \\
      0 & 1
    \end{array}
  \right]$, where rows show the words ``some'' and ``all'' and columns
  indicate worlds where {\sc some but not all} and {\sc all} are true,
  respectively, capturing the regularity that ``some'' can be said of
  situations where {\sc all} is actually the case. Assuming uniform
  costs for the two words, a rational speaker confronted with such a
  lexicon will prefer to say ``some'' only when they mean \textsc{some
    but not all}, and this is therefore the interpretation that a
  rational listener will place on ``some'' as well.

% To simulate this
%   situation, we place a Dirichlet(10, 10) prior on the meaning of
% ``some'' and a Dirichlet(1, 10) prior on the meaning of
% ``some'', giving speaker probabilities of
%   $\left[
%     \begin{array}{cc}
%       1.00 & 0.00 \\
%       0.00 & 1.00
%     \end{array}
%   \right]$
% and listener probabilities of
% $\left[
%     \begin{array}{cc}
%       .86 & .14 \\
%       .03 & .97
%     \end{array}
%   \right]$. These revised interpretation probabilities minimize
%   ambiguity (with weight concentrated on the diagonal), capturing the specificity implicature.

@@

Language learners and language users must consider word meanings both
within and across contexts. A critical part of this process is
reasoning pragmatically about agents' goals in individual
situations. In the current work we assume that agents communicating
with one another assume that there is a shared conventional lexicon
which they both rely on. They then
reason recursively about how this lexicon should be used to convey
particular meanings in context. This pair of assumptions allows us to
create a model that unifies two previously separate strands
of modeling work on language usage and acquisition, and account for a
variety of new phenomena. In particular, we consider new
explanations of disambiguation in early word learning and the acquisition of
scalar adjectives, and demonstrate that our model is capable of developing
novel and efficient communicative systems through iterated learning
within the context of a single simulated conversation.

Our work identifies two biases in communicators' behavior: a tendency for pragmatic speakers
and listeners to accentuate useful, sparse patterns in their
communicative systems (\S\ref{sec:emergence}), and for short,
high-frequency (``low cost'') expressions to
be assigned to common objects (\S\ref{sec:horn-emergence}).
Remarkably, while both of these biases are locally sub-optimal for individual agents,
they both systematically drive the overall communicative system
towards greater global efficiency. In the long term, these processes
should leave their mark on the structure of the language itself, which
may contribute to explaining how languages become 
optimized for effective communication \cite{zipf1936,piantadosi2011}.

>>>>

This resolves one paradox, but as soon as we attempt to add uncertainty about the meanings of words
to such a model, new paradoxes arise. Suppose the listener is a young
child who is uncertain about the lexicon their partner is using. The
obvious solution is for them to place a prior on the lexicon; they
then update their posterior based on whatever utterances and
contextual cues they observe, and in the mean time interpret each
utterance by making their best guess, marginalizing out this
uncertainty. This basic structure is captured in previous models of
Bayesian word learning \cite{frank2009}. But when combined with the
recursive pragmatic model, a new question arises: Given such a
listener, what model should the speaker use? A rational speaker
attempts to maximize the listener's likelihood of understanding, so if
an uncertain listener interpets by marginalizing over some posterior,
then a knowledgeable speaker should calculate this posterior and
marginalize themselves. But then the speaker is not using any
particular lexicon at all, and a rational learning listener should
therefore not be trying to infer the speaker's lexicon.\footnote{Of
  course, in reality both parties will generally have some
  uncertainty, making the situation even worse. If we start from an
  uncertain listener with a prior over lexicons, then a first-level
  uncertain speaker needs a prior over priors on lexicons, a
  second-level uncertain listener needs a prior over priors over
  priors, etc. The original $L_0 \rightarrow S_1 \rightarrow \dots$
  recursion was bad enough, but at least each step had a constant
  cost. This new recursion produces hyper$^n$-distributions for which
  inference almost immediately becomes intractable even in principle,
  since the dimensionality of the learning problem increases with each
  step. Yet, without this addition of new uncertainty at each level,
  the model would dissolve back into certainty as in the previous
  paragraph, making learning
  impossible.}
% In fact, such an uncertain listener's uncertainty dissolves; they
% may not know the lexicon, but they do
% know their posterior over the lexicon -- and because
% this is all that the speaker cares about, there is
% nothing for the listener to learn from the speaker's
% behavior.

@@

Formally, let an unadorned $L$ and $S$ denote our convention-based
listener and speaker. By assumption (c), the top-level listener has uncertainty over the lexicon; if the lexicon were known they would draw inferences as in $L_n$ above:
%\begin{align*}
%  P_L(\obj | \word) &= P_{S_{n-1}}(\word | \obj, \text{$L$'s data}) \times \prior \\
%    &= \sum_\lex P_{S_{n-1}}(\word | \obj, \lex) \times P(\lex | \text{$L$'s data}) \times \prior. \\
%    &= P_{L_n}(\obj | \word, \text{$L$'s data}). \\
%  P_S(\word | \obj) &\propto \exp\Big(\lambda \big(\log P_{L_n}(\obj | \word, \text{$S$'s data}) - \text{cost}(\word)\big)\Big).
%\end{align*}
%
\begin{align*}
  P_L(\obj | \word, \text{$L$'s data}) &=  \sum_\lex P_{L_n}(\obj | \word, \lex) P(\lex | \text{$L$'s data}) \\
\end{align*}
Here $L$'s data consists of $S$'s utterances together with whatever
other information $L$ may have about their meaning (e.g.~from
contextual clues). 
By assumption (b), $L$ treats these  utterances
as samples from the knowledgeable speaker $S_{n-1}$, not $S$, and thus
as being informative about the lexicon. For instance, when the data is a set of fully observed 
word-referent pairs $\{w_i, o_i\}$:
\begin{align*}
  P(\lex | \text{$L$'s data}) \propto P(\lex) \prod_i P_{S_{n-1}}(w_i|o_i,  \lex)
\end{align*}
The top-level speaker attempts to select the word which soft-maximizes
their utility, with utility now being defined in terms of the
informativity of the expectation (over lexicons) that the listener
will have for the right referent\footnote{An alternative model would
  have the speaker take the expectation over informativity, instead of
  the informativity of the expectation.}:
\begin{align*}
  P_S(\word | \obj, \text{$S$'s data}) &\propto \exp\Big(\lambda \big(\log  \sum_\lex P_{L_{n-1}}(\obj | \word, \lex) P(\lex|\text{$S$'s data}) - \text{cost}(\word)\big)\Big)
\end{align*}
Here $P(\lex|\text{$S$'s data})$ is defined similarly, when $S$ observes
$L$'s interpretations of various utterances, and treats them as
samples from $L_n$, not $L$.

@@

Many sets of words in natural language form scales in which each term
is more informative than, and is entailed by, the previous. ``Some'' and ``all'' form a scale of this
type. While ``I ate some of the cookies'' is compatible with the
followup ``in fact, I ate {\it all} of the cookies,'' the reverse is
not true. ``Might'' and ``must'' are another example, as are ``OK,''
``good,'' and ``excellent.'' All of these scales allow for {\it scalar
  implicatures} \cite{grice1975}: if a speaker uses a weaker term, a
listener is licensed to infer that the stronger term was not true in
this situation. So although ``I ate some of the cookies'' could in
principle be compatible with eating {\sc all} of them, the listener is lead
to believe that {\sc some-but-not-all} is the likely state of
affairs. We use the term {\it specificity implicature} here (following
\cite{bergen2012}) to denote that we are agnostic about whether such
implicatures are conventionalized or whether they use a
contextually-defined scale.

The recursive pragmatic reasoning portions of our model capture
findings on specificity implicature. Consider a representation of
scalar implicature as a matrix $\left[
    \begin{array}{cc}
      0.5 & 0.5 \\
      0 & 1
    \end{array}
  \right]$, where rows show the words ``some'' and ``all'' and columns
  indicate worlds where {\sc some but not all} and {\sc all} are true,
  respectively, capturing the regularity that ``some'' can be said of
  situations where {\sc all} is actually the case. Assuming uniform
  costs for the two words, a rational speaker confronted with such a
  lexicon will prefer to say ``some'' only when they mean \textsc{some
    but not all}, and this is therefore the interpretation that a
  rational listener will place on ``some'' as well. To simulate this
  situation, we place a Dirichlet(10, 10) prior on the meaning of
``some'' and a Dirichlet(1, 10) prior on the meaning of
``some'', giving speaker probabilities of
  $\left[
    \begin{array}{cc}
      1.00 & 0.00 \\
      0.00 & 1.00
    \end{array}
  \right]$
and listener probabilities of
$\left[
    \begin{array}{cc}
      .86 & .14 \\
      .03 & .97
    \end{array}
  \right]$. These revised interpretation probabilities minimize
  ambiguity (with weight concentrated on the diagonal), capturing the specificity implicature.

@@

If each agent in a multi-agent setting assumes the existence of a
shared, conventional norm underlying everyone's behaviour, and further
assumes that other agents are knowledgeable about this norm while they
themselves are not, then we are able to simultaneously preserve most
of the normative Bayesian framework, explicitly model learning and
recursive social reasoning within the same framework, and avoid
paradox. The resulting model unifies two previously separate strands
of modelling work on language usage and acquisition, and -- having
done so -- accounts for not just the phenomena these previous models
accounted for, but a variety of new ones as well. We propose new
explanations of the child language acquisition phenomena of
disambiguation without mapping \cite{horst2008} and the acquisition of
scalar adjectives (XX ask Mike what the levinson cite is really, see
also table), and demonstrate that our model is capable of developing
novel and efficient communicative systems through iterated learning
within the context of a single simulated conversation.

Accomplishing this requires limited deviations from rationality, and
as such necessarily introduces biases in agent behaviour. We've
identified two such biases here: the tendency for pragmatic speakers
and listeners to accentuate useful, sparse patterns in their
communicative systems (\S\ref{sec:emergence}), and for cheap words to
be assigned to common objects (\S\ref{sec:horn-emergence}).
Remarkably, while both are locally sub-optimal for individual agents,
they both systematically drive the overall communicative system
towards greater global efficiency. In the long term, these processes
should leave their mark on the structure of the language itself, which
may provide a partial explanation for how languages become so tightly
optimized for effective communication \cite{zipf1936,piantadosi2011}.
